An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2015-04-20 paper.tex
Final point in reponse to reviewer 1, point 2.

In order to use the kD-tree interpolation as a jump proposal in an
MCMC, we must be able to quickly find the neighborhood associated with
a given point to compute the jump probability (see
Eq.~\ref{eq:p-accept}).  This can be accomplished in $\order{\log N}$
time and constant space with the following algorithm, which is a
standard binary tree search.  Given the point, $\vtheta_i$ and the
tree, $T$:
\begin{enumerate}
\item If $T$ contains exactly one sample point, then its box is the
  associated neighborhood.  Otherwise:
\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
  contained in the ``left'' sub-tree, then return to Step 1,
  considering this sub-tree; otherwise return to Step 1, considering
  the ``right'' sub-tree.
\end{enumerate}

@@

The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $N_{crit}$ total samples. It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
The jump probability for this jump proposal is given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N V} , \ee
%
where $N$ is the total number of samples in the kD tree, and $V$ is the
volume of the intersection between the kD-tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
RJMCMC, it is important that this proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.

>>>>

In order to use the kD-tree interpolation as a jump proposal in an
MCMC, we randomly select a point stored in the kD tree with equal
probability, and propose from the associated neighbourhood.
Therefore, we must be able to quickly find the neighborhood associated
with a given point to compute the jump probability (see
Eq.~\ref{eq:p-accept}).  We introduce an additional parameter into the
neighbourhood search, $N_\mathrm{boxing}$, which describes the minimum
number of points in a kD box used as the neighbourhood for a point;
small $N_\mathrm{boxing}$ increases variance in the proposal but
resolves finer-scale structure in the posterior.  The neighbourhood
search can be accomplished in $\order{\log N}$ time and constant space
with the following algorithm, which is a modified binary tree search.
Given the randomly-chosen point, $\vtheta_i$, the tree, $T$, and
$N_\mathrm{boxing}$:
\begin{enumerate}
\item If $T$ contains fewer than $2N_\mathrm{boxing}$ points, then
  its box is the associated neighborhood.  Otherwise:
\item The tree $T$ has two sub-trees.  If the point $\vtheta_i$ is
  contained in the ``left'' sub-tree, then return to Step 1,
  considering this sub-tree; otherwise return to Step 1, considering
  the ``right'' sub-tree.
\end{enumerate}
The returned box is used for the jump proposal by drawing uniformly
from its interior, so the proposal density is 
\begin{equation}
  \label{eq:neighbourhood-jp}
  Q(\vtheta \to \vtheta^p) = \frac{N_\mathrm{box}}{N V},
\end{equation}
where $N$ is the number of points in the tree, $N_\mathrm{box}$ is the
number of points in the chosen kD box, and $V$ is the coordinate-space
volume of the box.

@@

The jump proposal chooses a stored sample at random and finds the largest kD-tree box containing it and fewer than $N_{crit}$ total samples. It then draws the covariance cell and proposes a new point uniformly from the intersection of the covariance cell with the original kD-tree box.
The jump probability for this jump proposal is still (c.f.\ Eq.\
\eqref{eq:neighbourhood-jp}) given by
%
\bel{modforward} P(\vtheta_i \rightarrow \vtheta_{i+1}) =
\frac{\Nbox}{N V} , \ee
%
where $N$ is the total number of samples in the kD tree, and $V$ is the
volume of the intersection between the kD-tree box and the correlation cell containing $\vtheta_{i+1}$. When used as the only jump proposal for an
RJMCMC, it is important that this proposal be capable of
proposing points in all allowed regions of parameter space; on the
other hand, when the kD proposal is used in a suite of other jump proposals, this is no longer a requirement.
